Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-01 analysis/paper/paper.Rmd
Changes in Methods and Introduction UP TO the "interaction" section.
saving here to make it easier to revert if unhappy with new intro focusing on hab diff with no frag x xlimate interaction).

---
title: "Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF]
      correspondence: true
  - Maria Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      intitute: [UF]
      correspondence: false

institute:
  - UF: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
editor_options:
  markdown:
    wrap: sentence
---

@@

<!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->

1.  Organisms simultaneously experience multiple forms of natural and anthropogenic disturbance.
    While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood.
    They could act in concert to exacerbate or dampen population dynamics.
    <!--# examples here? -->

2.  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. Of particular concern are the effects of fragmentation on species-rich tropical ecosystems. In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest \>1 km from an forest edge has decreased from 90% to only 75% [@haddadHabitatFragmentationIts2015].
    Fragmentation on its own can result in <!--# insert effects on plant populations -->

3.  In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stresses that are increasing with climate change (CITE IPCC?).
    For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014]. Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW).
In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Figure \@ref(fig:map)). These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried. In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990]; it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010].
While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others---including as *H. acuminata*---are usually found at lower densities and in the more shaded forest understory.
These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a].
In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002].
Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011].
Dispersed seeds (1--3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].
*Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*.
The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km.
Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments <!--# are we including 10ha fragments? --> were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->).
Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants.
We also surveyed plots regularly during the rainy season to identify any that flowered after the survey.
Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002].
Plants that were not found for three consecutive surveys were considered dead.
In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings <!--# are we using these data? -->.
There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Precipitation and potential evapotranspiration data were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought.
We used 3-month SPEI because *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Appendix \_\_**).

## Statistical Modeling

We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
We chose 36 months as a maximum lag because *H. acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
For example, the 1997 El Niño drought was one of the most severe droughts in the Amazon on record [@williamsonAmazonianTreeMortality2000] and correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also reported as exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008].
<!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->

@@

>>>>

---
title: "Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape"
author:
  - Eric R. Scott:
      email: scottericr@gmail.com
      institute: [UF_WEC]
      correspondence: true
  - Maria Uriarte:
      email: mu2126@columbia.edu
      institute: [Columbia]
      correspondence: false
  - Emilio M. Bruna:
      email: embruna@ufl.edu
      intitute: [UF_WEC,UF_LAS,BDFFP]
      correspondence: false

institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - Columbia: Department of Ecology, Evolution and Environmental Biology, Columbia University 1200 Amsterdam Avenue, New York, New York 10027 USA
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      number_sections: no
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
linkcolor: blue
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  forest fragmentation; vital rates; *Heliconia acuminata*; drought; ENSO; edge effects
highlights: |
  These are the highlights
editor_options:
  markdown:
    wrap: sentence
---

@@

<!-- Underemphasize the interaction between fragmentation and drought bc can't do the interaction in test. Instead it's "compare drought effects in fragmented and cf populations."  -->
<!-- In response to this I deleted #1; not looking at interaciton of multiple effects and not studying population dyanmics per se in this one. Suggest alluding to this in Discussion instead. -->

 <!-- [1]. Although the local extinction of species from fragments is common [1], the demographic mechanisms underlying these extinctions are rarely known [2]. However, it is often assumed – especially in the tropics – that these extinctions are driven by the dramatically altered abiotic conditions in fragments [3] reducing rates of reproduction or survivorship [4-6]. This is why understanding the way in which landscape structure and environmental conditions interact to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8].  -->

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat. Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*. However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, ultimately leading to local or regional population declines. This is thought to be particularly true in species-rich tropical forests, which continue to be cleared and fragmented at unprecedented rates. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].

<!-- CITATIONS_1: Haddad, others? -->
<!-- CITATIONS_2: Bruna et al J Veg Sci 2009, others? -->
<!-- CITATIONS_3: Bruna 1999 Nature, Laurance et al 1998 Ecology, Zartman 2015 Biotropica.  -->
<!-- CITATION_4: WRI Report? -->
<!-- CITATION_5: Brodie J et al 2012. TREE, Scheffers et al 2016. Science -->
<!-- CITATION_6: Zeng et al 2008 Environmental Research Letters, Mora et al 2013 Nature -->

Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*. These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] as well as increased drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*.
Severe droughts in the Amazon have been shown to reduce aboveground biomass through increased tree mortality and dieback [@williamsonAmazonianTreeMortality2000;[@phillipsDroughtSensitivityAmazon2009; @marengoDroughtAmazonia20052008; @lewis2010AmazonDrought2011].

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large areas of continuous forest, the BDFFP has forest fragment reserves isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and, in most cases, burning the downed trees once they dried (Figure \@ref(fig:map). In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

@@

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot distributed throughout Central Amazonia [@kressDiversityDistributionHeliconia1990] and is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in treefall gaps, others---such as *H. acuminata*---are found at lower densities in the darker and cooler forest understory. These species produce fewer infloresences and are pollinated by traplining rather than territorial hummingbirds [@berryHeliconiaIdentificationGuide1991a]. While some *Heliconia* species are at least partially self-compatible *[Kress Evolution 1983]*, pollen transfer experiments indicate self-fertilization in *Heliconia acuminata* is extremely rare *[darrigo unpubl. data]*.

In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@bruna2004]. Plants begin flowering at the start of the rainy season; reproductive plants have $\bar{x}$ = **1.1** flowering shoots (range = 1-7), each of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May, have 1--3 seeds per fruit ($\bar{x}$=2), and are eaten by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds germinate \~6 months after dispersal at the onset of the subsequent rainy season, with rates of germination and seedling establishment higher in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999].

@@

In 1997--1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *Heliconia acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m--41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table <!--# insert fig ref -->). This study uses data from 1-ha fragments and Continuous Forest sites.

Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify the annual surveys, during which we recorded the number of vegetative shoots each plant had, the height of each plant to the tallest leaf, and the identity of any flowering plants. We also surveyed plots regularly during the rainy season to identify any that flowered after the survey. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. Plants that were not found for three consecutive surveys were considered dead. There is a seven-fold range in *H. acuminata* density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because -- given its shallow roots and rhizome -- *H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales.
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Appendix \_\_**).

## Statistical Modeling of _______________ 
<!-- I think this is well described but could use some biologizing, eg,"This function allows the effect of a predictor on a response variable (e.g., SPEI on the growth rate of plants) to vary non-linearly through lag time and across predictor intensity (e.g, SPEI values from -2 to 2).  -->
We assessed the effects of drought history on plant vital rates using distributed lag non-linear models (DLNMs) [@gasparriniPenalizedFrameworkDistributed2017].
DLNMs model potentially delayed effects of predictor variables by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This function allows the effect of a predictor to vary non-linearly through lag time and across predictor intensity.
Crossbasis functions were created using the *dlnm* package with possible lags from 0--36 months [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020].
We chose 36 months as a maximum lag because prior *H. acuminata* individuals transplanted to forest fragments typically recovered from transplant shock within 36 months [@brunaExperimentalAssessmentHeliconia2002].
The crossbasis function was fit to the data in the context of a generalized additive model (GAM) using the *mgcv* package with restricted maximum likelihood [@woodGeneralizedAdditiveModels2017].
The general form of the vital rate models was as follows:

@@

The meteorological drought occurrences indicated by SPEI mostly agree with drought events in the Amazon reported in the literature.
For example, the drought associated with the 1997 El Niño Southern Oscillation (ENSO) event was one of the most severe on record for the Amazon [@williamsonAmazonianTreeMortality2000]; correspondingly, 1997 has the lowest SPEI values in our timeseries (Figure \@ref(fig:eda)d).\
The 2005 dry season (June--October) was also exceptionally dry, although this drought mostly affected the south western Amazon [@zengCausesImpacts20052008; @marengoDroughtAmazonia20052008]. <
!--# I know I'm missing some literature here.  Add more refs for more droughts. Any papers mentioning 2003 wet season drought? Also, Feb 2007?-->

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [7,8]. 
<!-- Didham RK, Kapos V, Ewers RM  2012. Rethinking the conceptual foundations of habitat fragmentation research. Oikos 121:161-170. -->
<!-- Driscoll DA, Banks SC, Barton PS, Lindenmayer DB, Smith AL  2013. Conceptual domain of the matrix in fragmented landscapes. Trends in Ecology & Evolution 28:605-613. -->

1. To set up the next paper: Organisms simultaneously experience multiple forms of disturbance. While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood. They could act in concert to exacerbate or dampen population dynamics.
